Inferring the effective start dates of non-pharmaceutical interventions during COVID-19 outbreaks

### 2020-05-24 ms/ms.tex
fig names and headings

% Table 1 %%%
\begin{table}[h]
\centering
\pgfplotstabletypeset[
    col sep=comma,
    string type,
    every head row/.style={before row=\hline,after row=\hline},
    every last row/.style={after row=\hline},
    ]{../data/NPI_dates.csv}
\caption{
\textbf{Official start of non-pharmaceutical interventions.}
The date of the first intervention is for a ban of public events, or encouragement of social distancing, or for school closures.
In all countries except Sweden, the date of the last intervention is for a lockdown. In Sweden, where a lockdown was not ordered during the studied dates, the last date is for school closures. Dates for European countries from \citet{Flaxman2020}, date for Wuhan, China from \citet{Pei2020}. See \autoref{fig:NPI_dates} for a visual presentation.
}
\label{table:NPI_dates}
\end{table}

%%% Fig NPI dates %%%
\begin{figure}[h]
    \centering
	\includegraphics[width=0.5\textwidth]{../figures/Fig-NPI_dates.pdf}
    \caption{
    \textbf{Official start of non-pharmaceutical interventions.}
	See \autoref{table:NPI_dates} for more details. Wuhan, China is not shown.
    } 
    \label{fig:NPI_dates}
\end{figure}

@@

%%% Fig 1 %%%
\begin{figure}[h]
    \centering
	\includegraphics[width=0.6\textwidth]{../figures/Fig1.pdf}
    \caption{
    \textbf{Official and effective start of non-pharmaceutical interventions.}
    	The difference between $\hat{\tau}$ the effective and $\tau^*$ the official start of NPI is shown for different regions. The effective NPI dates in Italy and Wuhan are significantly delayed compared to the official dates, whereas in Denmark, France, Spain, and Germany, the effective date is earlier than the official date.
	$\hat{\tau}$ is the posterior median, see \autoref{table:estimated-params}. $\tau^*$ is the last NPI date, see \autoref{table:NPI_dates}. Thin and bold lines show 95\% and 75\% credible intervals, respectively (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.)
    }
    \label{fig:fig1}
\end{figure}

@@

We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPI  significantly differs from the official date (\autoref{fig:fig1}), that is, the credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:fig1}). The exceptions are, as with the comparison to the simpler models, Austria, Belgium, and United Kingdom, as well another country (see below). Norway also has a relatively wide credible interval, which could be expected as it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
In the following, we describe our findings on late and early effective start of NPI in detail.

% Table WAIC %%%
\begin{table}[h]
\centering
\pgfplotstabletypeset[
    col sep=comma,
    string type,
    every head row/.style={before row=\hline,after row=\hline},
    every last row/.style={after row=\hline},
    ]{../figures/Table-WAIC.csv}
\caption{
\textbf{WAIC values for the different models.}
	WAIC (widely applicable information criterion)~\citep{gelman2013bayesian} values for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}. WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy.
}
\label{table:WAIC}
\end{table}

%%% Late
\paragraph*{Late effective start of NPIs.}
In both Wuhan, China, and in Italy we find that our estimated effective start of NPI $\hat{\tau}$ is significantly later than the official date $\tau^*$ (\autoref{fig:fig1}).

@@

%%% Fig Late %%%
\begin{figure}[h]
    \centering
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/posterior/Wuhan_τ_posterior.pdf}
    \end{subfigure}
  	~
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/posterior/Italy_τ_posterior.pdf}
    \end{subfigure}
    \caption{
	\textbf{Late effect of non-pharmaceutical interventions in Italy and Wuhan, China.}
    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples. Red line shows the official last NPI date $\tau^*$. Black line shows the estimated $\hat{\tau}$. Shaded area shows a 95\% credible interval (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$). 
    }
    \label{fig:late}
\end{figure}

%%% Early
\paragraph*{Early effective start of NPIs.}
In contrast, in some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:fig1}).
In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a  football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (2:1) with a crowd of 50,965 in Seville.
A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ on Mar~8 or~9 ($\pm$1.08 95\%CI), rather than Mar~14 (\autoref{fig:early}).

@@

%%% Fig 3 %%%
\begin{figure}[h]
    \centering
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/posterior/France_τ_posterior.pdf}
    \end{subfigure}
    ~
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/posterior/Spain_τ_posterior.pdf}
    \end{subfigure}
    \caption{
    \textbf{Early effect of non-pharmaceutical interventions in France and Spain.}
    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples. Red line shows the official last NPI date $\tau^*$. Black line shows the estimated $\hat{\tau}$. Shaded area shows a 95\% credible interval (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$). 
	}
	\label{fig:early}
\end{figure}

@@

%% Table 2 %%%
\begin{landscape}
\begin{table}[h]
\centering
\input{../figures/Table-estimated-params.tex}
\caption{
\textbf{Parameter estimates for different regions.}
See \autoref{eq:model} for model parameters.
All estimates are posterior medians.
75\% and 95\% credible intervals given for $\tau$, in days.
$\tau^*$ is the official last NPI date, see \autoref{table:NPI_dates}.
}
\label{table:estimated-params}
\end{table}
\end{landscape}

@@

%%% Fig 4 %%%
\begin{figure}[h]
    \centering
	\includegraphics[width=0.5\textwidth]{../figures/Fig4.pdf}
    \caption{
    \textbf{Late effective start of NPIs leads to under-estimation of daily confirmed cases.}
    Real number of daily cases in Italy in black (markers: data, line: time moving average). 
    Model posterior predictions are shown as colored lines (1,000 draws from the posterior distribution). 
    Shaded box illustrates a serial interval of seven days.
    \textbf{(A)} Using the official date $\tau^*$ for the start of the NPI,  the model under-estimates the number of cases seven days after the start of the NPI.
    \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
    Here, model parameters are estimates for Italy (\autoref{table:estimated-params}).
    } 
    \label{fig:fig4}
\end{figure}

@@

We have estimated the effective start date of NPIs in several geographical regions using an SEIR epidemiological model and an MCMC parameter estimation framework.
We find examples of both late and early effect of NPIs (\autoref{fig:fig1}).

>>>>

% Table - NPI dates %%%
\begin{table}[h]
\centering
\pgfplotstabletypeset[
    col sep=comma,
    string type,
    every head row/.style={before row=\hline,after row=\hline},
    every last row/.style={after row=\hline},
    ]{../data/NPI_dates.csv}
\caption{
\textbf{Official start of non-pharmaceutical interventions.}
The date of the first intervention is for a ban of public events, or encouragement of social distancing, or for school closures.
In all countries except Sweden, the date of the last intervention is for a lockdown. In Sweden, where a lockdown was not ordered during the studied dates, the last date is for school closures. Dates for European countries from \citet{Flaxman2020}, date for Wuhan, China from \citet{Pei2020}. See \autoref{fig:NPI_dates} for a visual presentation.
}
\label{table:NPI_dates}
\end{table}

%%% Fig - NPI dates %%%
\begin{figure}[h]
    \centering
	\includegraphics[width=0.5\textwidth]{../figures/Fig-NPI_dates.pdf}
    \caption{
    \textbf{Official start of non-pharmaceutical interventions.}
	See \autoref{table:NPI_dates} for more details. Wuhan, China is not shown.
    } 
    \label{fig:NPI_dates}
\end{figure}

@@

%%% Fig - tau summary %%%
\begin{figure}[h]
    \centering
	\includegraphics[width=0.6\textwidth]{../figures/Fig-tau-summary.pdf}
    \caption{
    \textbf{Official vs. effective start of non-pharmaceutical interventions.}
    	The difference between $\hat{\tau}$ the effective and $\tau^*$ the official start of NPI is shown for different regions. The effective NPI dates in Italy and Wuhan are significantly delayed compared to the official dates, whereas in Denmark, France, Spain, and Germany, the effective date is earlier than the official date.
	$\hat{\tau}$ is the posterior median, see \autoref{table:estimated-params}. $\tau^*$ is the last NPI date, see \autoref{table:NPI_dates}. Thin and bold lines show 95\% and 75\% credible intervals, respectively (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.)
    }
    \label{fig:tau-summary}
\end{figure}

@@

We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPI  significantly differs from the official date (\autoref{fig:tau-summary}), that is, the credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:tau-summary}). The exceptions are, as with the comparison to the simpler models, Austria, Belgium, and United Kingdom, as well another country (see below). Norway also has a relatively wide credible interval, which could be expected as it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
In the following, we describe our findings on late and early effective start of NPI in detail.

% Table - WAIC %%%
\begin{table}[h]
\centering
\pgfplotstabletypeset[
    col sep=comma,
    string type,
    every head row/.style={before row=\hline,after row=\hline},
    every last row/.style={after row=\hline},
    ]{../figures/Table-WAIC.csv}
\caption{
\textbf{WAIC values for the different models.}
	WAIC (widely applicable information criterion)~\citep{gelman2013bayesian} values for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}. WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy.
}
\label{table:WAIC}
\end{table}

%%% Late
\paragraph*{Late effective start of NPIs.}
In both Wuhan, China, and in Italy we find that our estimated effective start of NPI $\hat{\tau}$ is significantly later than the official date $\tau^*$ (\autoref{fig:tau-summary}).

@@

%%% Fig - Late %%%
\begin{figure}[h]
    \centering
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/posterior/Wuhan_τ_posterior.pdf}
    \end{subfigure}
  	~
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/posterior/Italy_τ_posterior.pdf}
    \end{subfigure}
    \caption{
	\textbf{Late effect of non-pharmaceutical interventions in Italy and Wuhan, China.}
    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples. Red line shows the official last NPI date $\tau^*$. Black line shows the estimated $\hat{\tau}$. Shaded area shows a 95\% credible interval (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$). 
    }
    \label{fig:late}
\end{figure}

%%% Early
\paragraph*{Early effective start of NPIs.}
In contrast, in some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:tau-summary}).
In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a  football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (2:1) with a crowd of 50,965 in Seville.
A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ on Mar~8 or~9 ($\pm$1.08 95\%CI), rather than Mar~14 (\autoref{fig:early}).

@@

%%% Fig - Early %%%
\begin{figure}[h]
    \centering
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/posterior/France_τ_posterior.pdf}
    \end{subfigure}
    ~
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/posterior/Spain_τ_posterior.pdf}
    \end{subfigure}
    \caption{
    \textbf{Early effect of non-pharmaceutical interventions in France and Spain.}
    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples. Red line shows the official last NPI date $\tau^*$. Black line shows the estimated $\hat{\tau}$. Shaded area shows a 95\% credible interval (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$). 
	}
	\label{fig:early}
\end{figure}

@@

%% Table - estimated parameters %%%
\begin{landscape}
\begin{table}[h]
\centering
\input{../figures/Table-estimated-params.tex}
\caption{
\textbf{Parameter estimates for different regions.}
See \autoref{eq:model} for model parameters.
All estimates are posterior medians.
75\% and 95\% credible intervals given for $\tau$, in days.
$\tau^*$ is the official last NPI date, see \autoref{table:NPI_dates}.
}
\label{table:estimated-params}
\end{table}
\end{landscape}

@@

%%% Fig - Prediction %%%
\begin{figure}[h]
    \centering
	\includegraphics[width=0.5\textwidth]{../figures/Fig-prediction.pdf}
    \caption{
    \textbf{Late effective start of NPIs leads to under-estimation of daily confirmed cases.}
    Real number of daily cases in Italy in black (markers: data, line: time moving average). 
    Model posterior predictions are shown as coloured lines (1,000 draws from the posterior distribution). 
    Shaded box illustrates a serial interval of seven days.
    \textbf{(A)} Using the official date $\tau^*$ for the start of the NPI,  the model under-estimates the number of cases seven days after the start of the NPI.
    \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
    Here, model parameters are estimates for Italy (\autoref{table:estimated-params}).
    } 
    \label{fig:fig-prediction}
\end{figure}

@@

We have estimated the effective start date of NPIs in several geographical regions using an SEIR epidemiological model and an MCMC parameter estimation framework.
We find examples of both late and early effect of NPIs (\autoref{fig:tau-summary}).
